---
title: "Lab4"
output: html_document
---

### C
The data campy.dat contain the number of cases of campylobacter infections in the north of the province Quebec (Canada) in four week intervals from January 1990 to the end of October 2000. It has 13 observations per year and 140 observations in total. Assume that the number of infections $c_t$ at each time point follows an independent Poisson distribution when conditioned on a latent AR(1)-process $x_t$, that is
$$c_t|x_t \sim Poisson(exp(x_t))$$
 
where $x_t$ is an AR(1)-process as in a). Implement and estimate the model in Stan, using suitable priors of your choice. Produce a plot that contains both the data and the posterior mean and 95% credible intervals for the latent intensity $\theta_t = exp(x_t)$ over time. [Hint: Should $x_t$ be seen as data or parameters?]

```{r}
library(rstan)
campy = read.table("/Users/oskarhiden/Git/TDDE07 Bayesian Learning/Lab 4/campy.dat", header = TRUE)
campy = as.list(campy)

StanModel = '
data{
  int<lower=0> N;
  int<lower=0> y[N];
}
parameters {
  real mu;
  real phi;
  real<lower=0> sigma2;
  real x[N];
}
model {
  mu ~ normal(2,50); // Normal with mean 2(from log10), st.dev. 50
  sigma2 ~ scaled_inv_chi_square(1,2); // Scaled-inv-chi2 with nu 1, sigma 10
  phi ~ uniform(-1, 1);
  for (n in 2:N){
    x[n] ~ normal(mu + phi * x[n-1], sqrt(sigma2));
    y[n] ~ poisson(exp(x[n]));
  }
}
'
T_2 = 140
campy = campy$c
data = list(N=T_2, y=campy)
burnin = 1000
niter = 2000
fit_campy = stan(model_code=StanModel,data=data,
              warmup=burnin,iter=niter,chains=4)

post_summary = summary(fit_campy)
mean = post_summary$summary[,1]
low = post_summary$summary[,4]
high = post_summary$summary[,8]

plot(campy)
points(exp(mean[4:(length(mean)-1)]), col="red", type="l")
points(exp(low[4:(length(mean)-1)]), col="blue", type="l")
points(exp(high[4:(length(mean)-1)]), col="blue", type="l")
```

### D
Now, assume that we have a prior belief that the true underlying intensity $\theta_t$ varies more smoothly than the data suggests. Change the prior for $\sigma^2$ so that it becomes informative about that the AR(1)-process increments $\epsilon_t$ should be small. Re-estimate the model using Stan with the new prior and produce the same plot as in c). Has the posterior for $\theta_t$ changed?

```{r}
StanModel_inform = '
data{
  int<lower=0> N;
  int<lower=0> y[N];
}
parameters {
  real mu;
  real phi;
  real<lower=0> sigma2;
  real x[N];
}
model {
  mu ~ normal(2,50); // Normal with mean 2(from log10), st.dev. 50
  sigma2 ~ scaled_inv_chi_square(300,0.1); // Scaled-inv-chi2 with nu 1, sigma 10
  phi ~ uniform(-1, 1);
  for (n in 2:N){
    x[n] ~ normal(mu + phi * x[n-1], sqrt(sigma2));
    y[n] ~ poisson(exp(x[n]));
  }
}
'

fit_campy = stan(model_code=StanModel_inform,data=data,
                 warmup=burnin,iter=niter,chains=4)

print(fit_campy, digits_summary = 3)

post_summary = summary(fit_campy)
mean = post_summary$summary[,1]
low = post_summary$summary[,4]
high = post_summary$summary[,8]

plot(campy)
points(exp(mean[4:(length(mean)-1)]), col="red", type="l")
points(exp(low[4:(length(mean)-1)]), col="blue", type="l")
points(exp(high[4:(length(mean)-1)]), col="blue", type="l")
```

The posterior for $\theta_t$ has changed, the standard deviation of $\theta_t$ has decreased.